\large{\bf Looking forwards and backwards through locally regulated populations }

### 2022-09-15 forwards-and-backwards.tex
various things

\begin{proof}[of Lemma~\ref{lem:SecondMomentMass}]
    Using equation~\eqref{mgp3} with $\phi = 1$,
    \begin{align*}
        \IE[\langle 1, \eta_t \rangle^2]
        &=
        \IE[\langle 1, \eta_0 \rangle^2]
        +
        \IE\left[
        \int_0^t
            2 \langle 1, \eta_s \rangle
            \langle F(x, \eta_s), \eta_s(dx) \rangle
            +
            \frac{\theta}{N} \langle 2 + F(x, \eta_s)/\theta, \eta_s(dx) \rangle
        ds
        \right] .
    \end{align*}
    Since $F(m)1_{m > 0}$ is bounded above and $\theta/N$ converges,
    we have for an appropriate $C$ that
    \begin{align*}
        \IE[\langle 1, \eta_t \rangle^2]
        \le
        C + C \int_0^t \IE[\langle 1, \eta_s \rangle] ds
        + C \int_0^t \IE[\langle 1, \eta_s \rangle^2] ds .
    \end{align*}
    The result follows from Lemma~\ref{Lemma:UniformlyBoundedMass} and Gronwall's inequality.
\end{proof}

@@

\begin{proof}
    We can explicitly write the quadratic variation of $M^x_t$ as
    \[
        [M^x]_t
        =
        \frac{\theta}{N} \int_0^t \langle
            \int_{\IR^d} \psi_{t-s}^{\epsilon,x}(z)^2 q_\theta(y,dz)
            +
            \left(1 + \frac{F(y, \eta_s)}{\theta}\right) \psi_{t-s}^{\epsilon,x}(y)^2,
        \eta_s(dy) \rangle  ds .
    \]
    Lemma~\ref{PsiBoundHS} implies that $\psi_{t-s}^{\epsilon,x}(y) \le C \epsilon^{-d}$.
    Using this and the fact that $F(x)$ is bounded above,
    \begin{align} \label{QuadVarBound}
    [M^x]_t
        &\leq
        \frac{C \theta}{N \epsilon^d} \int_0^t \left\{
            \langle \int \psi_{t-s}^{\epsilon,x}(z) q_\theta(y,dz), \eta_s(dy) \rangle
            +
            \langle \psi_{t-s}^{\epsilon,x}(y), \eta_s(dy)\rangle
        \right\} ds 
    \end{align}
    We can now take expectations and use Corollary~\ref{cor:BoundLocalDensity} to prove the result.
\end{proof}

@@

    Let $f \in C_c^2$ (that is, compactly supported and with smooth second derivative).
    We will show convergence of $\int f(x) \rho_\epsilon * \eta_t(x) dx$ (which is enough to establish weak convergence of the measures $\rho_\epsilon * \eta_t(x) dx$).
    \comment{Alison writes: no it's not weak - this would be vague convergence. Take $f \in C^2$ to mean bounded first and second derivatives, which is weak convergence.}
    For this note that
    \begin{align*}
        \int f(x) \rho_\epsilon * \eta_t(x) dx
        &=
        % \int \int f(x) \rho_\epsilon (x-y) \eta(dy) dx
        % \\ & =
        % \langle \int f(x) \rho_\epsilon(x-y) dx, \eta(dy) \rangle
        % \\ &=
        \langle \rho_\epsilon f(y), \eta_t(dy) \rangle .
    \end{align*}
    Substituting this into the martingale problem for $\eta_t$, \eqref{mgp1},
    we get
    \begin{align}
         \label{MGPT_epsilon}
    \langle \rho_{\epsilon} * f, \eta_t \rangle
        & = \nonumber 
        \langle \rho_{\epsilon} * f, \eta_0 \rangle
        +
        \theta \int_0^t
            \langle
                \int (\rho_{\epsilon} * f(z) - \rho_{\epsilon} * f(y)) q_\theta(y, dz) ,
                \eta_s(dy)
            \rangle
        ds
        \\ & {}
        +
        \int_0^t \langle \rho_\epsilon * f(y) F(y, \eta_s), \eta_s(dy) \rangle ds
        + \widehat{M}_t,
    \end{align}
    with $\widehat{M}_t$ a martingale. Now note that
    by the Intermediate Value Theorem,
    \begin{align}
        \label{ShortBoundLaplacianHS}
        \left| \theta \int (\rho_{\epsilon} * f(z) - \rho_{\epsilon} * f(y)) q_\theta(y,z) dz \right|
        &=
        \left| \theta( p_{\epsilon^2 + 1/\theta} * f(y) - p_{\epsilon^2} * f(y) ) \right| \nonumber
        \\ &=
        \left| \Delta p_{\epsilon+1/\tilde{\theta}} * f(y) \right| \nonumber
        \\ &=
        \left| p_{\epsilon^2 + 1/\tilde{\theta}} * \Delta f(y) \right|
        \leq
        \| \Delta f \|_{\infty}
    \end{align}
    for some $\tilde{\theta} \in [0,1/\theta]$.
    Proceeding as in the proof of Lemma~\ref{lem:eta_f_bound},
    we can use that $F(x)$ is bounded above and Gronwall's inequality to conclude that
    \[
        \IE\left[\langle \rho_{\epsilon} * f(y) F(y, \eta_t), \eta_t(dy) \rangle \right]
        \text{ is uniformly bounded for } t \in [0,T].
    \]

    The quadratic variation of $\widehat{M}_t$ is
    \[
        [ \widehat{M} ]_t
        =
        \frac{\theta}{N} \int_0^t \langle
            p_{1/\theta}*(p_{\epsilon^2} * f(\cdot)^2)(y)
            +
            \left( 1 + \frac{F(\rho_\epsilon*\eta_s(y))}{\theta} \right)
            p_{\epsilon^2} * f(y)^2
            ,
            \eta_s(dy)
        \rangle ds ,
    \]
    so using boundedness of $f$ and $F$,
    \begin{align*}
        \IE\left[ [\widehat{M}]_t \right]
        \leq
        \frac{\theta}{N} \| f \|_\infty C
        \int_0^t \mathbb{E}\left[ \langle 1, \eta_s(dy) \right] ds
        <
        C \frac{\theta}{N} t ,
    \end{align*}
    where in the last inequality we have used Lemma~\ref{Lemma:UniformlyBoundedMass}.
    Hence, using the Burkholder-Davis-Gundy and Jensen's inequalities we can conclude that
    \[
        \IE[\sup_{0 \leq t \leq T} |\widehat{M}_t| ]
        \le
        C
        \IE\left[ [\widehat{M}]_t \right]^{1/2}
        \le
        \sqrt{ C \frac{\theta}{N} T } .
    \]
    Therefore, by taking supremum and expectation and in \eqref{MGPT_epsilon}
    we conclude that
    \begin{equation}
        \label{eq:boundedIntegralAgainstf}
        \IE[ \sup_{0 \leq t \leq T} \langle T_{\epsilon^2} f, \eta_t \rangle ]
        \text{ is uniformly bounded (for fixed } f \text{)}
    \end{equation}
    To conclude $C$-tightness we show that, for $T>0$, and $\alpha > 0$
    there is some $\delta >0$ so that for $\theta$ and $N$ large enough,
    \begin{align}
        &
        \IP\left[
            \sup_{s,t \in [0,T], |s-t| < \delta}
            \left|
                \int f(x) \rho_\epsilon*\eta_s(x) dx
                -
                \int f(x) \rho_\epsilon*\eta_t(x) dx
            \right|
            >
            \alpha
        \right]
        \\ &=
        \IP\left[
            \sup_{s,t \in [0,T], |s-t| < \delta}
            \left|
                \langle p_{\epsilon^2} * f, \eta_s \rangle
                - \langle p_{\epsilon^2} * f, \eta_t \rangle
            \right|
            >
            \alpha
        \right]
        <
        \alpha \label{conditionCtightness}
    \end{align}
    But, from our estimates above, we have
    \begin{align*}
        \left|
        \langle p_{\epsilon^2} * f, \eta_s \rangle
        -
        \langle p_{\epsilon^2} * f, \eta_t \rangle
        \right|
        \leq
        ( \| \Delta f \|_\infty + C_F \|f\|_\infty) |u-v|
        \sup_{0 \leq u \leq T } \langle 1, \eta_u \rangle
        +
        \sup_{u \leq T} |\widehat{M}_u| .
    \end{align*}
    Taking expectations we get,
    \begin{equation}
        \IE\left[ \left|
            \langle T_{\epsilon^2} f, \eta_s \rangle
            -
            \langle T_{\epsilon^2} f, \eta_t \rangle
        \right| \right]
        \leq
        \widehat{C} |s-t|
        +
        \widehat{C} \left( \frac{\theta}{N} T \right)^{1/2} .
    \end{equation}
    Using this and Markov's inequality give that \eqref{conditionCtightness}
    holds with $\delta = \alpha/(3 \widehat{C}|s-t|)$,
    for $N$, $\theta$ big enough so $\widehat{C}\sqrt{T\theta/N}$ is bounded by $\alpha/3$.
    (Note the last condition is possible by \eqref{ConditionsForConvergence}.)
    Thus we have concluded $C$-tightness. 
\end{proof}

\begin{remark} \label{remark:BoundedIntegral}
Similar as we got equation (\ref{eq:boundedIntegralAgainstf}), from equation \ref{MGPT_epsilon} we can conclude that
\[ \IE\left[\sup_{0 \leq t \leq T} \int \rho_{\epsilon/2} * \eta_t(x) dx \right] = C < \infty. \]
By using this and the same pattern used to prove Lemma~\ref{lem:SecondMomentMass}, we can then conclude that,
\[ \IE\left[  \sup_{0 \leq t \leq T} \int \left(\rho_{\epsilon/2} * \eta_t(x) \right)^2 dx \right] = C < \infty. \]
Since,
\begin{align*}
\langle \rho_\epsilon * \eta_t(z), \eta(dz) \rangle & = \sum_{y \in \eta_t} \sum_{z \in \eta_t} \rho_\epsilon(y-z) \eta(z) \eta(y) \\ &= \int  \sum_{y \in \eta_t} \sum_{z \in \eta_t} \rho_{\epsilon/2}(y-x) \rho_{\epsilon/2}(z-x) \eta(z) \eta(y) \\ &= \int (\rho_{\epsilon/2}*\eta_t(x))^2 dx.
\end{align*}
This allows us to conclude that $\IE[\langle \rho_\epsilon*\eta_t(z),\eta_t(dz) \rangle ] < \infty$. This bound will be useful in the next section.
\end{remark}

>>>>

\begin{proof}[of Lemma~\ref{lem:SecondMomentMass}]
    Using equation~\eqref{mgp3} with $\phi = 1$,
    \begin{align*}
        \IE[\langle 1, \eta_t \rangle^2]
        &=
        \IE[\langle 1, \eta_0 \rangle^2]
        +
        \IE\left[
        \int_0^t
            2 \langle 1, \eta_s \rangle
            \langle F(x, \eta_s), \eta_s(dx) \rangle
            +
            \frac{\theta}{N} \langle 2 + F(x, \eta_s)/\theta, \eta_s(dx) \rangle
        ds
        \right] .
    \end{align*}
    Since $F$ is bounded above and $\theta/N$ converges,
    we have for an appropriate $C$ that
    \begin{align*}
        \IE[\langle 1, \eta_t \rangle^2]
        \le
        C + C \int_0^t \IE[\langle 1, \eta_s \rangle] ds
        + C \int_0^t \IE[\langle 1, \eta_s \rangle^2] ds .
    \end{align*}
    The result follows from Lemma~\ref{Lemma:UniformlyBoundedMass} and Gronwall's inequality.
\end{proof}

@@

\begin{proof}
    We can explicitly write the quadratic variation of $M^x_t$ as
    \[
        [M^x]_t
        =
        \frac{\theta}{N} \int_0^t \langle
            \int_{\IR^d} \psi_{t-s}^{\epsilon,x}(z)^2 q_\theta(y,dz)
            +
            \left(1 + \frac{F(y, \eta_s)}{\theta}\right) \psi_{t-s}^{\epsilon,x}(y)^2,
        \eta_s(dy) \rangle  ds .
    \]
    Lemma~\ref{PsiBoundHS} implies that $\psi_{t-s}^{\epsilon,x}(y) \le C \epsilon^{-d}$.
    Using this and the fact that $F(x)$ is bounded above,
    \begin{align} \label{QuadVarBound}
    [M^x]_t
        &\leq
        \frac{C \theta}{N \epsilon^d} \int_0^t \left\{
            \langle \int \psi_{t-s}^{\epsilon,x}(z) q_\theta(y,dz)
            +
            \psi_{t-s}^{\epsilon,x}(y)
            , \eta_s(dy)\rangle
        \right\} ds 
    \end{align}
    We can now take expectations and use Corollary~\ref{cor:BoundLocalDensity} to prove the result.
\end{proof}

@@

    Let $f \in C_c^2$ (that is, compactly supported and with smooth second derivative).
    We will show convergence of $\int f(x) \rho_\epsilon * \eta_t(x) dx$ (which is enough to establish weak convergence of the measures $\rho_\epsilon * \eta_t(x) dx$).
    \comment{Alison writes: no it's not weak - this would be vague convergence. Take $f \in C^2$ to mean bounded first and second derivatives, which is weak convergence.}
    For this note that
    \begin{align*}
        \int f(x) \rho_\epsilon * \eta_t(x) dx
        &=
        % \int \int f(x) \rho_\epsilon (x-y) \eta(dy) dx
        % \\ & =
        % \langle \int f(x) \rho_\epsilon(x-y) dx, \eta(dy) \rangle
        % \\ &=
        \langle \rho_\epsilon * f(y), \eta_t(dy) \rangle .
    \end{align*}
    Substituting this into the martingale problem for $\eta_t$, \eqref{mgp1},
    we get
    \begin{align}
         \label{MGPT_epsilon}
    \langle \rho_{\epsilon} * f, \eta_t \rangle
        & = \nonumber 
        \langle \rho_{\epsilon} * f, \eta_0 \rangle
        +
        \theta \int_0^t
            \langle
                \int (\rho_{\epsilon} * f(z) - \rho_{\epsilon} * f(y)) q_\theta(y, dz) ,
                \eta_s(dy)
            \rangle
        ds
        \\ & {}
        +
        \int_0^t \langle \rho_\epsilon * f(y) F(y, \eta_s), \eta_s(dy) \rangle ds
        + \widehat{M}_t,
    \end{align}
    with $\widehat{M}_t$ a martingale. Now note that
    by the Intermediate Value Theorem,
    \begin{align}
        \label{ShortBoundLaplacianHS}
        \left| \theta \int (\rho_{\epsilon} * f(z) - \rho_{\epsilon} * f(y)) q_\theta(y,z) dz \right|
        &=
        \left| \theta( p_{\epsilon^2 + 1/\theta} * f(y) - p_{\epsilon^2} * f(y) ) \right| \nonumber
        \\ &=
        \left| \Delta p_{\epsilon+1/\tilde{\theta}} * f(y) \right| \nonumber
        \\ &=
        \left| p_{\epsilon^2 + 1/\tilde{\theta}} * \Delta f(y) \right|
        \leq
        \| \Delta f \|_{\infty}
    \end{align}
    for some $\tilde{\theta} \in [0,1/\theta]$.
    Proceeding as in the proof of Lemma~\ref{lem:eta_f_bound},
    we can take expectations of equation~\eqref{MGPT_epsilon},
    use \eqref{ShortBoundLaplacianHS} and the fact that $F(x)$ is bounded above,
    and then Gronwall's inequality to conclude that
    \[
        \IE\left[\langle \rho_{\epsilon} * f(y), \eta_t(dy) \rangle \right]
        \text{ is uniformly bounded for } t \in [0,T].
    \]

    The quadratic variation of $\widehat{M}_t$ is
    \[
        [ \widehat{M} ]_t
        =
        \frac{\theta}{N} \int_0^t \langle
            p_{1/\theta}*(p_{\epsilon^2} * f(\cdot)^2)(y)
            +
            \left( 1 + \frac{F(\rho_\epsilon*\eta_s(y))}{\theta} \right)
            p_{\epsilon^2} * f(y)^2
            ,
            \eta_s(dy)
        \rangle ds ,
    \]
    so using boundedness of $f$ and $F$,
    \begin{align*}
        \IE\left[ [\widehat{M}]_t \right]
        \leq
        \frac{\theta}{N} \| f \|_\infty (1 + C_F/\theta)
        \int_0^t \mathbb{E}\left[ \langle 1, \eta_s(dy) \right] ds
        <
        C \frac{\theta}{N} t ,
    \end{align*}
    where in the last inequality we have used Lemma~\ref{Lemma:UniformlyBoundedMass}.
    Hence, using the Burkholder-Davis-Gundy inequality and then Jensen's inequality we can conclude that
    \[
        \IE[\sup_{0 \leq t \leq T} |\widehat{M}_t| ]
        \le
        C
        \IE\left[ [\widehat{M}]_T \right]^{1/2}
        \le
        \sqrt{ C \frac{\theta}{N} T } .
    \]
    Therefore, by taking supremum and expectation and in \eqref{MGPT_epsilon}
    we conclude that
    \begin{equation}
        \label{eq:boundedIntegralAgainstf}
        \IE[ \sup_{0 \leq t \leq T} \langle p_{\epsilon^2} * f, \eta_t \rangle ]
        \text{ is uniformly bounded (for fixed } f \text{)} .
    \end{equation}
    To conclude $C$-tightness we show that, for $T>0$, and $\alpha > 0$
    there is some $\delta >0$ so that for $\theta$ and $N$ large enough,
    \begin{align}
        % &
        % \IP\left[
        %     \sup_{s,t \in [0,T], |s-t| < \delta}
        %     \left|
        %         \int f(x) \rho_\epsilon*\eta_s(x) dx
        %         -
        %         \int f(x) \rho_\epsilon*\eta_t(x) dx
        %     \right|
        %     >
        %     \alpha
        % \right]
        % \\ &=
        \IP\left[
            \sup_{s,t \in [0,T], |s-t| < \delta}
            \left|
                \langle \rho_{\epsilon} * f, \eta_s \rangle
                - \langle \rho_{\epsilon} * f, \eta_t \rangle
            \right|
            >
            \alpha
        \right]
        <
        \alpha \label{conditionCtightness}
    \end{align}
    Now, we can use \eqref{MGPT_epsilon} to write this difference
    as an integral, then bound the terms as above to get
    \begin{align*}
        \left|
        \langle \rho_{\epsilon} * f, \eta_s \rangle
        -
        \langle \rho_{\epsilon} * f, \eta_t \rangle
        \right|
        \leq
        ( \| \Delta f \|_\infty + C_F \|f\|_\infty) |s - t|
        \sup_{0 \leq u \leq T } \langle 1, \eta_u \rangle
        +
        \sup_{u \leq T} |\widehat{M}_u| .
    \end{align*}
    Taking expectations we get,
    \begin{equation}
        \IE\left[ \left|
            \langle \rho_\epsilon * f, \eta_s \rangle
            -
            \langle \rho_\epsilon * f, \eta_t \rangle
        \right| \right]
        \leq
        \widehat{C} |s-t|
        +
        \widehat{C} \left( \frac{\theta}{N} T \right)^{1/2} .
    \end{equation}
    Using this and Markov's inequality give that \eqref{conditionCtightness}
    holds with $\delta = \alpha/(3 \widehat{C}|s-t|)$,
    for $N$, $\theta$ big enough so $\widehat{C}\sqrt{T\theta/N}$ is bounded by $\alpha/3$.
    (Note the last condition is possible by \eqref{ConditionsForConvergence}.)
    Thus we have concluded $C$-tightness. 
\end{proof}

\comment{
    Peter writes: the following Remark refers to the proof of Lemma~\ref{lem:SecondMomentMass},
    which was previously wrong and is now fairly different.
    However, I don't see how to prove the Remark using any similar method,
    because to bound $\int p_{\epsilon^2 / 2} * \eta_t(x)^2 dx$
    it seems like we need control on $\rho_\epsilon * \eta_t(x)$ uniformly in $x$
    (and, integrability), which I don't think we have?
    But, I'm probably being dense.
    See after this for an attempted rewrite.
}

\begin{remark} \label{remark:BoundedIntegral}
    Similar as we got equation (\ref{eq:boundedIntegralAgainstf}), from equation \ref{MGPT_epsilon} we can conclude there exists a $C < \infty$ such that
    \[
        \IE\left[\sup_{0 \leq t \leq T} \int p_{\epsilon^2/2} * \eta_t(x) dx \right] < C .
    \]
    By using this and the same pattern used to prove Lemma~\ref{lem:SecondMomentMass},
    we can then conclude there is a (different) $C < \infty$ such that
    \[
        \IE\left[  \sup_{0 \leq t \leq T} \int \left( p_{\epsilon^2/2} * \eta_t(x) \right)^2 dx \right]
        <
        C .
    \]
    Then, because
    \begin{align*}
        \langle p_{\epsilon^2} * \eta_t(x), \eta(dx) \rangle
        &=
        \langle p_{\epsilon^2/2} * p_{\epsilon^2/2} * \eta_t(x), \eta(dx) \rangle
        \\ &=
        \int_{\IR^d} \left( p_{\epsilon^2/2} * \eta_t(x) \right)^2 dx .
    \end{align*}
    we can conclude that
    $\IE[\langle \rho_\epsilon*\eta_t(x), \eta_t(dx) \rangle ]$ is uniformly bounded,
    which will be useful in the next section.
\end{remark}

\comment{Here's my start at rewriting that Remark;
    however, I'm stuck since $\theta(p_{\epsilon^2 + 1/\theta} - p_{\epsilon^2}) * \eta$
    isn't bounded by $p_{\epsilon^2} * \eta$ (uniformly in $\epsilon$, anyhow).
}

\begin{remark} \label{remark:BoundedIntegral2}
    Below, we will want to bound
    \[
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        =
        \int_{\IR^d} \int_{\IR^d} p_{\epsilon^2}(x, y) \eta_t(dx) \eta_t(dy) .
    \]
    Writing $H(\eta) = \langle \rho_\epsilon * \eta, \eta \rangle$,
    note that since $p_{\epsilon^2}$ is symmetric,
    $H(\eta + \delta_z / N) = H(\eta) + 2 \rho_\epsilon * \eta(z) / N + p_{\epsilon^2}(z, z) / N^2$,
    so following Section~\ref{sec:heuristics},
    \begin{align*}
        \langle \rho_\epsilon * \eta_t, \eta_t \rangle
        &=
        \langle \rho_\epsilon * \eta_0, \eta_0 \rangle
        +
        2 \theta \int_0^t \left \langle \int_{\IR^d}
            \left(
               \rho_\epsilon * \eta_s(y)
               -
               \rho_\epsilon * \eta_s(x)
            \right)
            q_\theta(x, dy)
        , \eta_s(dy) \right \rangle ds
        \\ & \qquad {}
        +
        2 \int_0^t
        \langle
            \rho_\epsilon * \eta_s(x) F(x, \eta_s)
        , \eta_s(dx) \rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right)
        + M_t ,
    \end{align*}
    where $M_t$ is a martingale with quadratic variation
    \begin{align*}
        [M]_t
        &=
        2 \frac{\theta}{N} \int_0^t
        \left \langle
            \int_{\IR^d}
                \rho_\epsilon * \eta_s(y)^2
            q_\theta(x, dy)
        , \eta_s(dx) \right\rangle
        ds
        \\ & \qquad {}
        +
        2 \int_0^t
        \left \langle
            \rho_\epsilon * \eta_s(x)^2 (1 + F(x, \eta_s) / \theta) 
        , \eta_s(dx) \right\rangle
        ds
        +
        O\left( \frac{1}{\epsilon^d N^2} \right) .
    \end{align*}
\end{remark}
